Domain Specific Languages and Code Synthesis using Haskell

### 2014-04-04 paper.tex
Fixing the paper

A Domain Specific Language (DSL) is a special purpose language,
designed to encapsulate possible computations in a specific
domain. Following our earlier examples of MATLAB, SQL,
%% earlier examples of?
Verilog, and spreadsheets, the domains would be scientific modeling,
database queries and updates, hardware circuits, and financial computations, respectively.
Considering SQL specifically, there is nothing SQL does that could not
be done in Java or C, or any other general purpose programming
language. SQL simply bundles the actions needed to
interact with a database into a useable and productive package,
%% is useable the correct adjective here?  Java and C are useable languages too.
and the language becomes the interface to communicate requests
to the database engine.
To pick a concrete example,
consider the act of trying to list a dynamically updated leader-board
for an online programming contest:
\begin{Code}
SELECT ROUND(SUM(s.Score)) as ss, t.TeamName FROM Solution s -- an aggregate score and team name
   LEFT JOIN Team t ON SolutionTeam = TeamId                 -- where the team has the correct id
   GROUP BY s.SolutionTeam                                   -- grouped by team name
   ORDER BY ss DESC                                          -- ordered by score
   LIMIT 20                                                  -- returning the first 20
\end{Code}
In a handful on lines, this query performs a complex
database search, and some computation, finding
the names and aggregate scores of the top 20 teams.

There are two fundamental types of DSLs
(Figure~\ref{fig:types-of-dsls}).
%
The first, like the SQL example, is when a DSL is a first class language,
with its own compiler or interpreter, and is often used in
its own ecosystem. (Figure \ref{fig:types-of-dsls}(1)) 
All the examples mentioned so far fall
into this category. The primary difference between SQL and,
for instance, Java is one of scope and focus, though sometimes
DSLs grow to be as large as general purpose languages.
%
The other class of DSLs are languages
embedded in a host language(Figure \ref{fig:types-of-dsls}(2)). 
Such languages
can have the look and feel of being their own languages,
but leverage the host language's existing
ecosystem and initial semantics. 
(Figure \ref{fig:types-of-dsls}(2))
It is this second
class of DSLs we are interested in, and is the subject of
this paper.

@@

Polymorphic values, which are analogous to the type \verb|Object| in Java,
or \verb|void*| pointers in C, are expressed using lower-case letters.
These polymorphic values can have constraints expressed over them,
using the Haskell equivalent of an object hierarchy.
Finally, a Haskell function that takes a list, and returns a list
is written as using an arrow: \verb|[a] -> [a]|.

@@

\end{Code}
This function sorted a list, using a variant of quicksort in which the pivot is
the first element of the list.
\begin{itemize}
\item The first line is the type for \verb|sort|. This is $\forall$\verb|a|, such that
\verb|a| can be \verb|Ord|ered (admits comparisons like \verb|<=|), the function
takes and return a list of such \verb|a|'s.
\item The second line says that an empty list is already sorted.
\item The remaining lines state that a (non-empty) list can be
sorted by taking the first and rest of the list (called \verb|x| and \verb|xs|, respectively),
and sorting the values before this pivot and after this pivot,
and concatenating theses intermediate values together.
\item Finally, intermediate values can be named using the \verb|where| syntax;
in this case the values of \verb|before| and \verb|after|.
\end{itemize}

@@

It is possible to give specific types extra powers, like equality and comparison,
using the class based overloading system. In the case of \verb|Maybe|, we
can give the \verb|Maybe| type the ability to test for equality, using an
instance.
\begin{Code}
instance Eq a => Eq (Maybe a) where        
  Just a  == Just b  = a == b
  Nothing == Nothing = True
  _       == _       = False
\end{Code}

This states that for any type that can be tested for equality,
we can also check \verb|Maybe| at the same type. We take the \verb|Maybe|
apart, using pattern matching on \verb|Just|, to check the internal value.a

In Haskell, Side-effects such as writing to the screen
or reading the keyboard are described using a \verb|do|-notation, for example:

@@

An EDSL is a library in a host language which has the look, feel and semantics of its own language,
customized to a specific problem domain.
EDSLs, by reusing the facilities and tools of the host language, considerably lower the cost of developing and maintaining a DSL.
Benefiting from Haskell's concise syntax, the Haskell community, and the functional programming
community in general, have taken the ideas
of EDSLs and developed a large number of DSLs
which provide higher-level interfaces and abstractions for well-understood systems.
We now look at two examples of DSLs, one for automatically generating test cases for software testing,
and a second for specifying hardware circuit behaviors.

@@

\verb|prop_reverse_twice| is a regular Haskell function,
that takes a list of Int, and returns a boolean, based
on the validity of what is being proposed, specifically
that is a two reverses cancel each other out. 
Here is the neat part --- \verb|prop_reverse_twice| is {\bf also\/} a
domain specific statement, and as such can also be considered
a sub-language inside of Haskell itself.  This style
of using functions, in this case functions with names
prefixed with \verb|prop_|, taking a number of typed
arguments, and returning a conditional is a small language.
The property written in Haskell is also an Embedded Domain Specific
Language for properties.
We can run our EDSL, using a function called \verb|quickCheck|:
\begin{Code}
Prelude Test.QuickCheck> quickCheck prop_reverse_twice
+++ OK, passed 100 tests.
\end{Code}
By running \verb|quickCheck| with our explicit and specific property,
we execute our ESDL inside Haskell. The \verb|quickCheck| function
generates 100 test cases for our property, and executes
them on the fly. If they all hold, then the system prints
a message reflecting this. The test cases are generated using
the type class system -- QuickCheck gives specific types the
power of test-case generation -- and the \verb|quickCheck| function
uses this to generate random tests.

@@

To take another example, consider describing hardware.
Hardware description languages and functional languages have
long enjoyed a fruitful partnership and
{\bf Lava} is the name given to a class of Haskell DSLs
that implement a function-based version of the hardware description
language Ruby~\cite{Jones:90:Ruby,Hutton:93:RubyInterp}. Ruby, not to be confused with the
modern programming language with the same name, was based
on relations, not functions, and was inspired by
the seminal work in $\mu$FP~\cite{Sheeran:84:muFP}.

@@

We can simulate sequential circuits with the
same directness as the combinational functions we invoked.
\begin{Code}
GHCi> toSeq (cycle [True,False,False])
T : F : F : T : F : F : T : F : F : ...
GHCi> counter low (toSeq (cycle [True,False,False]))
1 : 1 : 1 : 2 : 2 : 2 : 3 : 3 : 3 : ...
\end{Code}

@@

This level of circuit specification has been used to great
effect in many Lava and Lava-like languages. One notable
instance is Hawk, a Lava-like EDSL which was used to specify
the entire micro-architecture of the Pentium-Pro, including the
super-scaler design, and register bypass capabilities.

If DSLs are so powerful as an idiom for library design, then why have DSLs not taken over?
As a means for expressing things that can be {\em simulated\/}, EDSLs are an invaluable
design pattern. 
However, not everything is a simulation. What if we wanted to use an EDSL to express
something that we want to run somewhere else, not inside the Haskell system?
Can we use Lava to generate circuits, and run them on FPGAs? Can we use EDSLs
to generate code for embedded processors, or GPUs? Such an ability, to 
synthesis external solutions, would be extremely useful. We can extend
the EDSL idiom to do so, with significant caveats. The remainder of this
paper is about how we capture and offshore work from inside an EDSL, what
this capability can be used for, and what the limitations are.

@@

EDSLs are simply a way of thinking about a library
of provided functions, often called combinators, because they combine
their arguments into terms inside the DSL language and types. In our
Lava example above, the \verb|register| combinator takes an initial
value, and an incoming stream of values, and gives the new stream,
delayed by a single cycle, with the initial value occupying the initial
cycle. Critically, \verb|register| is compositional; it combines
smaller parts of the DSL to make larger solutions. If a DSL
follows this composability carefully by design, an important alternative implementation
is possible.

The most common flavor of DSL is one  that uses a so-called shallow embedding~(Figure~\ref{fig:types-of-dsls}(2a)), where values are computed with directly.
The result of a computation in a shallow DSL is a value. All the examples so far are shallow.
But there is another class of DSLs, 
specifically {\em DSLs that use a deep embedding\/} build an abstract syntax tree~(Figure~\ref{fig:types-of-dsls}(2b)).
The result of a computation inside a Deeply Embedded DSL (Deep EDSL)
is a structure, not a value, and this structure can be used to compute a value,
or be cross-compiled before being evaluated~\cite{Elliott:03:CompileDSEL-JFP}. Such ``Deep'' EDSLs
follow the composability mantra pedantically, by design and mandate.

Historically, EDSLs have been shallow; simply a way of structuring an API for a library. 
However, Deep EDSLs have the ability to {\em stage\/} code, that is executing a program
can generate another program, much like the well-known YACC DSL,
but at the cost of significantly restricting what forms of the DSL can
generate valid output from.
There is a growing number of Deep EDSLs, and a body of research around their
form and limitations.
The unifying theme is that Deep EDSLs can be pragmatic, productive and useful.

In this section, we will investigate the basic structure of Deep EDSL compared to Shallow EDSLs,
and look at three pragmatic tricks for improving the usefulness of Deep EDSLs.

@@

A deeply embedded DSL is a DSL that exposes its own structure.
Rather than functions operating directly on values (a shallow DSL),
a deep DSL builds a structure, then allows some secondary agent to
provide interpretation of this structure.
To make this idea concrete, consider a DSL for arithmetic,
with addition, subtraction, multiplication, and constants.
We want to be able to {\bf run\/} this DSL. For a shallow
embedding this is trivial - we just use the built in 
arithmetic. A deep embedding is where things get
interesting. Consider a data-type for our arithmetic.

\begin{Code}
data E where
 Lit :: Integer -> E
 Add :: E -> E -> E
 Sub :: E -> E -> E
 Mul :: E -> E -> E
 deriving Eq
\end{Code}

Now we overload the arithmetic to use this E datatype;
in Haskell Num is our overloading for integral arithmetic.

\begin{Code}
instance Num E where
  fromInteger n = Lit n
  e1 + e2 = Add e1 e2
  e1 - e2 = Sub e1 e2
  e1 * e2 = Mul e1 e2
\end{Code}

If we build expressions of type E, we can
observe the structure of the computation.
\begin{Code}
GHCi> 1 + 2 * 3
Add (Lit 1) (Mul (Lit 2) (Lit 3))
\end{Code}

This is profound, and the key idea that makes deep embeddings work. We can write
an expression and extract a tree of {\em what\/} to do, not a direct result.
With deep embeddings, it is common to also
write a run function, that computes the result
of a captured computation.
\begin{Code}
run :: E -> Int
run (Lit n)   = n
run (Add a b) = run a + run b
run (Sub a b) = run a - run b
run (Mul a b) = run a * run b
\end{Code}

@@

In order to make deep DSLs practical, there are two folklore tricks that are almost
always used.
The first trick allows the capture of functions, via dummy arguments. The second trick can observe loops,
via some form of observable sharing. 
%And finally, we call observe imperative statements, by using normalization. 
%We will overview both these tricks, before concluding with shortcomings, extensions, and remaining challenges.

@@

f :: E -> E
f x = x + 1

@@

Here we have a function that acts over our new type
\verb|E|, and returns a new \verb|E|. How can we capture
this function?
The trick is to invent a unique E, and pass it as a (dummy) argument
to \verb|f|.

\begin{Code}
data E where
  Lit :: Integer -> E
  Add :: E -> E -> E
  Sub :: E -> E -> E
  Mul :: E -> E -> E
  Var :: String -> E    -- new constructor
\end{Code}

@@

\begin{Code}
-- Just running the function
GHCi> f 4
Add (Lit 4) (Lit 1)
-- reifing the function, using our unique E ``Var''.
GHCi> f (Var "x")
Add (Var "x") (Lit 1)   -- reified version of the function
\end{Code}

@@

There are many places this design pattern can be used. One example is the specification
of surface textures as functions; it is possible to export these into code executable
on GPUs, simultaneously lifting the abstractions used to write textures, and speeding
up how fast a implementation of the same operations would run.  There is nothing
that is specific about Haskell, or even functional languages here. Indeed, the same
ideas have been used in Java, for a VHDL generator\cite{Bellows:98:JHDL}. 
Haskell, with its powerful abstractions,
allows deep DSLs to almost feel like a straightforward shallow embedding.

\subsection{How to spot a loop}

@@

@@

\begin{figure}[!t]
  \centering
   \begin{minipage}{0.6\textwidth}
     \centering
\footnotesize\begin{Code}[fontsize=\tiny]
entity counter is
  port(rst : in std_logic;
       clk : in std_logic;
       clk_en : in std_logic;
       restart : in std_logic;
       inc : in std_logic;
       output : out std_logic_vector(3 downto 0));
end entity counter;
architecture str of counter is
  signal sig_2_o0 : std_logic_vector(3 downto 0);
  ...
begin
  sig_2_o0 <= sig_5_o0 when (inc = '1')  else sig_6_o0;
  sig_5_o0 <= std_logic_vector(...);
  sig_6_o0 <= "0000" when (restart = '1') else sig_10_o0;
  sig_10_o0_next <= sig_2_o0;
  proc14 : process(rst,clk) is
  begin
    if rst = '1' then
      sig_10_o0 <= "0000";
    elsif rising_edge(clk) then
      if (clk_en = '1') then
        sig_10_o0 <= sig_10_o0_next;
  ....
end architecture;
\end{Code}
  \end{minipage}
  \caption{VHDL Generated by Kansas Lava for {\tt counter}}
  \label{fig:counter-VHDL}
%  \vspace{-0.1in}
\end{figure}

@@

\noindent
There are also extensions to the basic techniques. The principal ones are:
\begin{itemize}
\item Internal function calls can captured as notes on a graph, 
rather than directly inlined~\cite{Mainland:10:Nikola}. 
This helps compilation of large programs, giving a basic separate compilation capability.
\item \verb|do|-statement can be reified, by normalization~\cite{Sculthorpe:13:ConstrainedMonad,Svenningsson:13:CompositionalMonadReification,Persson:11:GenericMonadicConstructs}. This result, called monadic
reification, is surprising. There are strong technical reasons to be believe monadic reification
should be impossible. However the normalization refactors the constraints that, by themselves, would
be impossible to solve, and matches them up, 1-on-1, with a matching witness, allowing the whole
\verb|do|-notation to be solved and reified.
Monadic reification is a recent discovery, but has already been used in several Deep DSLs,
including Feldspar\cite{Axelsson:2010:DandI-Feldspar} and Sunroof\cite{Bracker:14:Sunroof}.
\item Control flow is problematic, and cannot be used directly. However,
there is a generalization of Haskell boolean that does allow deep embedding capture~\cite{hackage:boolean}.
Using this library, a DSL with control flow can be constructed, but it needs to be explicit
code, at the DSL level, using constructors. The usage is clumsy, but workable, and we should
be able to do better.
\end{itemize}

So where does this leave Deep DSLs? They are clearly a useful tool for the
language implementor, but come with costs and limitations. How can we therefore
push the state of the art, and allow more of the Haskell language to reified?
There are two primary shortcoming.
One we have discussed already: control flow and pattern matching remain a thorn in Deep DSLs.

Parametric polymorphism, one of the strengths of functional program, is the other
issue for deep DSLs. We need a specific structure to represent what we have captured,
and arbitrary polymorphism interferes with this. Current systems sidestep this
by always instantiating at a specific type, but this is expensive because the size of
the captured program can expand exponentially.
Polymorphism was the technical reason that it was thought that monadic reification
would not work, but in that case it was side-stepped by normalization; this technique
does not generalize to all polymorphism.

Deep DSLs is a value-level way of extracting an expression, but there are others.
Quasi-quoting is another, at the syntactic level. Haskell comes with an extensive
template system, called Template Haskell, that is used for DSLs.
Perhaps the future of Deep DSLs is some hybrid between expression generation
and quasi-quoting, combining the best of both systems?

>>>>

A Domain Specific Language (DSL) is a special purpose language,
designed to encapsulate possible computations in a specific
domain. Following our earlier examples of MATLAB, SQL,
%% earlier examples of?
Verilog, and spreadsheets, the domains would be scientific modeling,
database queries and updates, hardware circuits, and financial computations, respectively.
Considering SQL specifically, there is nothing SQL does that could not
be done in Java or C, or any other general purpose programming
language. SQL simply bundles the actions needed to
interact with a database into a useable and productive package,
%% is useable the correct adjective here?  Java and C are useable languages too.
and the language becomes the interface to communicate requests
to the database engine.

There are two fundamental types of DSLs
(Figure~\ref{fig:types-of-dsls}).
%
The first is when a DSL is a first class language,
with its own compiler or interpreter, and is often used in
its own ecosystem. (Figure \ref{fig:types-of-dsls}(1)) 
All the examples mentioned so far fall
into this category. The primary difference between the SQL DSL and,
for instance, Java is one of scope and focus, though sometimes
DSLs grow to be as large as general purpose languages.
%
The other class of DSLs are languages
embedded in a host language (Figure \ref{fig:types-of-dsls}(2)). 
Such languages
can have the look and feel of being their own language,
but leverage the host language's existing
ecosystem and initial semantics. 
(Figure \ref{fig:types-of-dsls}(2))
It is this second
class of DSLs we are interested in, and is the subject of
this paper.

@@

Polymorphic values are expressed using lower-case letters.
and play a similar role 
\verb|void*| pointers in C
and polymorphic arguments in the Java generics facility.
These polymorphic values can have constraints expressed over them,
using the Haskell equivalent of an object hierarchy.

Finally, a function is written using an arrow from 
argument type to result type.
So, in Haskell, a function that takes a list and returns a list
is written as using: \verb|[a] -> [a]|.

@@

\end{Code}
This function sorted a list, using a variant of quicksort in which the pivot is
the first element of the list.
\begin{itemize}
\item The first line is the type for \verb|sort|. This is $\forall$\verb|a|, such that
\verb|a| can be \verb|Ord|ered (admits comparisons like \verb|<=|), the function
takes and return a list of such \verb|a|'s.
\item The second line says that an empty list is already sorted.
\item The remaining lines state that a non-empty list can be
sorted by taking the first and rest of the list (called \verb|x| and \verb|xs|, respectively),
and sorting the values before this pivot and after this pivot,
and concatenating theses intermediate values together.
\item Finally, intermediate values can be named using the \verb|where| syntax;
in this case the values of \verb|before| and \verb|after|.
\end{itemize}

@@

It is possible to give specific types extra powers, like equality and comparison,
using the class based overloading system. In the case of \verb|Maybe|, we
can give the \verb|Maybe| type the ability to test for equality, using an
instance.
\begin{Code}
instance Eq a => Eq (Maybe a) where        
  Just a  == Just b  = a == b
  Nothing == Nothing = True
  X       == X       = False
\end{Code}

This states that for any type that can be tested for equality,
we can also check \verb|Maybe| at the same type. We take the \verb|Maybe|
apart, using pattern matching on \verb|Just|, to check the internal value.

In Haskell, side-effects such as writing to the screen
or reading the keyboard are described using a \verb|do|-notation, for example:

@@

An EDSL is a library in a host language which has the look, feel and semantics of its own language,
customized to a specific problem domain.
EDSLs, by reusing the facilities and tools of the host language, considerably lower the cost of both
developing and maintaining a DSL.
Benefiting from Haskell's concise syntax, the Haskell community, and the functional programming
community in general, have taken the ideas
of EDSLs and developed a large number of DSLs
which provide higher-level interfaces and abstractions for well-understood systems.
We now look at two examples of DSLs, one for automatically generating test cases for software testing,
and a second for specifying hardware circuit behaviors.

@@

\verb|prop_reverse_twice| is a regular Haskell function,
that takes a list of Int, and returns a boolean, based
on the validity of what is being proposed, specifically
that is a two reverses cancel each other out. 
Here is the neat part --- \verb|prop_reverse_twice| is {\bf also\/} a
domain specific statement, and as such can also be considered
a sub-language inside of Haskell itself.  This style
of using functions, in this case functions with names
prefixed with \verb|prop_|, taking a number of typed
arguments, and returning a conditional is a small language.
The property written in Haskell is also an Embedded Domain Specific
Language for properties, called QuickCheck~\cite{Claessen:00:QuickCheck}.
We can run our EDSL, using a function also called \verb|quickCheck|:
\begin{Code}
Prelude Test.QuickCheck> quickCheck prop_reverse_twice
+++ OK, passed 100 tests.
\end{Code}
By running \verb|quickCheck| with our explicit and specific property,
we execute our ESDL inside Haskell. The \verb|quickCheck| function
generates 100 test cases for our property, and executes
them on the fly. If they all hold, then the system prints
a message reflecting this. The test cases are generated using
the type class system -- QuickCheck gives specific types the
power of test-case generation -- and the \verb|quickCheck| function
uses this to generate random tests.

@@

To take another example, consider describing hardware.
Hardware description languages and functional languages have
long enjoyed a fruitful partnership and
{\bf Lava} is the name given to a class of Haskell DSLs
that implement a function-based version of the hardware description
language Ruby~\cite{Jones:90:Ruby,Hutton:93:RubyInterp}. Ruby, not to be confused with the
modern programming language with the same name, was based
on relations, which was in turn inspired by
the seminal work in $\mu$FP~\cite{Sheeran:84:muFP}.

@@

We can simulate sequential circuits with the
same directness as the combinational functions we invoked.
\begin{Code}
GHCi> toSeq (cycle [True,False,False])
True : False : False : True : False : False : True : False : False : ...
GHCi> counter low (toSeq (cycle [True,False,False]))
1 : 1 : 1 : 2 : 2 : 2 : 3 : 3 : 3 : ...
\end{Code}

@@

This level of circuit specification has been used to great
effect in many Lava and Lava-like languages. One notable
instance is Hawk~\cite{Matthews:1998:Hawk}, a Lava-like EDSL which was used to specify
the entire micro-architecture of the Pentium-Pro, including the
super-scaler design, and register bypass capabilities.

Now,if DSLs are so powerful as an idiom for library design, then why have DSLs not taken over?
As a means for expressing things that can be {\em simulated\/}, EDSLs are an invaluable
design pattern. 
However, not everything is a simulation. What if we wanted to use an EDSL to express
something that we want to run somewhere else, not inside the Haskell system?
Can we use Lava to generate circuits, and run them on FPGAs? Can we use EDSLs
to generate code for embedded processors, or GPUs? Such an ability, to 
synthesis external solutions, would be extremely useful. We can extend
the EDSL idiom to do so, with significant caveats. The remainder of this
paper is about how we capture and offshore work from inside an EDSL, what
this capability can be used for, and what the limitations are.

@@

EDSLs are simply a way of thinking about a library
of provided functions, often called combinators, because they combine
their arguments into terms inside the DSL language. In our
Lava example above, the \verb|register| combinator takes an initial
value, and an incoming stream of values, and gives the new stream,
delayed by a single cycle, with the initial value occupying the initial
cycle. Critically, \verb|register| is compositional; it combines
smaller parts of the DSL to make larger solutions. If a DSL
follows this composability carefully by design, an important alternative implementation
is possible.

The most common flavor of DSL is one  that uses a so-called shallow embedding~(Figure~\ref{fig:types-of-dsls}(2a)), where values are computed with directly.
The result of a computation in a shallow DSL is a value. All the examples so far are shallow.
But there is another class of DSLs, 
specifically {\em DSLs that use a deep embedding\/} build an abstract syntax tree~(Figure~\ref{fig:types-of-dsls}(2b)).
The result of a computation inside a deeply Embedded DSL (deep EDSL)
is a structure, not a value, and this structure can be used to compute a value,
or be cross-compiled before being evaluated~\cite{Elliott:03:CompileDSEL-JFP}. Such ``deep'' EDSLs
follow the composability mantra pedantically, by design and mandate.

Historically, EDSLs have been shallow; simply a way of structuring an API for a library. 
However, deep EDSLs have the ability to {\em stage\/} code, that is executing a program
can generate another program, much like the well-known YACC DSL,
but at the cost of significantly restricting what forms of the DSL can
generate valid output from.
There is a growing number of deep EDSLs, and a body of research around their
form and limitations.
The unifying theme is that deep EDSLs can be pragmatic, productive and useful.

In this section, we will investigate the basic structure of deep EDSL compared to Shallow EDSLs,
and look at three pragmatic tricks for improving the usefulness of deep EDSLs.

@@

A deeply embedded DSL is a DSL that exposes its own composition and structure.
Rather than functions operating directly on values (a shallow DSL),
a deep DSL builds a structure, then allows some secondary agent to
provide interpretation of this structure.
To make this idea concrete, consider a DSL for arithmetic,
with addition, subtraction, multiplication, and constants.
We want to be able to {\bf run\/} this DSL. For a shallow
embedding this is trivial - we just use the built in 
arithmetic. A deep embedding is where things get
interesting. Consider a data-type for our arithmetic.

\begin{Code}
data Expr where
 Lit :: Integer -> Expr
 Add :: Expr -> Expr -> Expr
 Sub :: Expr -> Expr -> Expr
 Mul :: Expr -> Expr -> Expr
 deriving Eq
\end{Code}

Now we overload the arithmetic to use this E datatype;
in Haskell \verb|Num| is our overloading for integral arithmetic.

\begin{Code}
instance Num Expr where
  fromInteger n = Lit n
  e1 + e2 = Add e1 e2
  e1 - e2 = Sub e1 e2
  e1 * e2 = Mul e1 e2
\end{Code}

If we build expressions of type E, we can
observe the structure of the computation.
\begin{Code}
GHCi> 1 + 2 * 3 :: Expr
Add (Lit 1) (Mul (Lit 2) (Lit 3))
\end{Code}

This is profound, and the key idea that makes deep embeddings work. We can write
an expression and extract a tree of {\em what\/} to do, not a direct result.
With deep embeddings, it is common to also
write a run function, that computes the result
of a captured computation.
\begin{Code}
run :: Expr -> Integer
run (Lit n)   = n
run (Add a b) = run a + run b
run (Sub a b) = run a - run b
run (Mul a b) = run a * run b
\end{Code}

@@

In order to make deep DSLs practical, there are two additional tricks in the DSL folklore that are almost
always used.
The first trick allows the capture of functions, via dummy arguments. The second trick can observe loops,
via some form of observable sharing. 
%And finally, we call observe imperative statements, by using normalization. 
%We will overview both these tricks, before concluding with shortcomings, extensions, and remaining challenges.

@@

f :: Expr -> Expr -> Expr
f x = x + 1

@@

Here we have a function that acts over our new type
\verb|Expr|, and returns a new \verb|Expr|. How can we capture
this function?
The trick is to invent a unique \verb|Expr|, and pass it as a (dummy) argument
to \verb|f|.

\begin{Code}
data Expr where
  Lit :: Integer -> Expr
  Add :: Expr -> Expr -> Expr
  Sub :: Expr -> Expr -> Expr
  Mul :: Expr -> Expr -> Expr
  Var :: String -> Expr    -- new constructor
\end{Code}

@@

\begin{Code}
-- Just running the function
GHCi> f 4
Add (Lit 4) (Lit 1)
-- reifing the function, using our unique Var.
GHCi> f (Var "x")
Add (Var "x") (Lit 1)   -- reified version of the function
\end{Code}

@@

This idea scales to multi-argument functions. Consider \verb|g|:

g :: Expr -> Expr -> Expr
g x y = x * x + y + 2

Given two argument protoypical arguments to \verb|g| will capture the function.

\begin{Code}
GHCi> g (Var "x") (Var "y")
Add (Add (Mul (Var "x") (Var "x")) (Var "y")) (Lit 2)
\end{Code}

There are many places this design pattern can be used. One example is the specification
of surface textures as functions; it is possible to export these into code executable
on GPUs, simultaneously lifting the abstractions used to write textures, and speeding
up how fast an implementation of the same operations would run.  There is nothing
that is specific about Haskell, or even functional languages here. Indeed, the same
ideas have been used in Java, for a VHDL generator\cite{Bellows:98:JHDL}. 
Haskell, with its powerful abstractions,
allows deep DSLs to almost feel like a straightforward shallow embedding.

\subsection{How to Spot a Loop}

@@

register :: a -> Signal a -> Signal a
register d s = Register d s

@@

\begin{figure}[!t]
  \centering
   \begin{minipage}{0.6\textwidth}
     \centering
\footnotesize\begin{Code}[fontsize=\small]
entity counter is
  port(rst : in std_logic;
       clk : in std_logic;
       clk_en : in std_logic;
       restart : in std_logic;
       inc : in std_logic;
       output : out std_logic_vector(3 downto 0));
end entity counter;
architecture str of counter is
  signal sig_2_o0 : std_logic_vector(3 downto 0);
  ...
begin
  sig_2_o0 <= sig_5_o0 when (inc = '1')  else sig_6_o0;
  sig_5_o0 <= std_logic_vector(...);
  sig_6_o0 <= "0000" when (restart = '1') else sig_10_o0;
  sig_10_o0_next <= sig_2_o0;
  proc14 : process(rst,clk) is
  begin
    if rst = '1' then
      sig_10_o0 <= "0000";
    elsif rising_edge(clk) then
      if (clk_en = '1') then
        sig_10_o0 <= sig_10_o0_next;
  ....
end architecture;
\end{Code}
  \end{minipage}
  \caption{VHDL Generated by Kansas Lava for {\tt counter}}
  \label{fig:counter-VHDL}
%  \vspace{-0.1in}
\end{figure}

@@

\noindent
There are also extensions to the basic techniques. The principal ones are:
\begin{itemize}
\item Internal function calls can captured as notes on a graph, 
rather than directly inlined~\cite{Mainland:10:Nikola}. 
This helps compilation of large programs, giving a basic separate compilation capability.
\item \verb|do|-statement can be reified, by normalization~\cite{Sculthorpe:13:ConstrainedMonad,Svenningsson:13:CompositionalMonadReification,Persson:11:GenericMonadicConstructs}. This result, called monadic
reification, is surprising. There are strong technical reasons to be believe monadic reification
should be impossible. However the normalization refactors the constraints that, by themselves, would
be impossible to solve, and matches them up, 1-on-1, with a matching witness, allowing the whole
\verb|do|-notation to be solved and reified.
Monadic reification is a recent discovery, but has already been used in several Deep DSLs,
including Feldspar\cite{Axelsson:2010:DandI-Feldspar} and Sunroof\cite{Bracker:14:Sunroof}.
\item Control flow is problematic, and cannot be used directly. However,
there is a generalization of Haskell boolean that does allow deep embedding capture~\cite{hackage:boolean}.
Using this library, a DSL with control flow can be constructed, but it needs to be explicit
code, at the DSL level, using constructors. 
The \verb|mux2| function used previously is a simplification of this idea.
The usage is clumsy, but workable, and we should be able to do better.
\end{itemize}

So where does this leave Deep DSLs? They are clearly a useful design pattern for the
language implementor, but come with costs and limitations. How can we therefore
push the state of the art, and allow more of the Haskell language to reified?
There are two primary shortcoming.
One we have discussed already: control flow and pattern matching remain a thorn in Deep DSLs.

Parametric polymorphism, one of the strengths of functional program, is the other
issue for deep DSLs. We need a specific structure to represent what we have captured,
and arbitrary polymorphism interferes with this. Current systems sidestep this
by always instantiating at a specific type, but this is expensive because the size of
the captured program can expand exponentially.
Polymorphism was the technical reason that it was thought that monadic reification
was not possible, but in that case it was side-stepped by normalization; this technique
does not generalize to all polymorphism.

Deep DSLs is a value-level way of extracting an expression, but there are ways of
extracting expressions.
Quasi-quoting is another mechanism for extracting expressions, but at the syntactic level. 
Haskell comes with an extensive template system, called Template Haskell\~cite{Sheard:02:metahaskell}, 
that is often used for DSLs. There is a sense of unease with such solutions, however, much
in the same way the C pre-processor is used, but not considered elegant. The principal issue
is that the syntax of Haskell is huge, perhaps 100 syntactical terms. An expression-based
solution, such as a deep embedding, can avoid the need to re-write front translations.
Quasi-quoting has one important advantage, specifically it can cope with control flow,
and deconstruction of values. So perhaps the future of deep DSLs is some hybrid between
expression generation and quasi-quoting, combining the best of both systems.
